---
title: "Study of The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Luis A Sandino"
output: pdf_document
---

***
```{r,echo=FALSE,message=FALSE,results='hide'}
library(lubridate)
library(dplyr)
library(ggplot2)
library(knitr)
Sys.setlocale("LC_TIME", "English")
options(scipen = 999)
```

Created on `r now()` using `r R.version$version.string`

***

## Synopsis

This project performs a basic inferential analysis of the ToothGrowth data in the R datasets package. According to the help file: *The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).*

##1. Load the ToothGrowth data and perform some basic exploratory data analyses

Load the dataset and convert variable dose to factor:
```{r,tidy=TRUE}
# 
library(datasets)
data(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```

Inspect the structure of the data:
```{r,tidy=TRUE}
str(ToothGrowth)
```

##2. Provide a basic summary of the data

Obtain a summary of the data:
```{r,tidy=TRUE}
ToothGrowth %>% 
        group_by(supp,dose) %>%
        summarise(q25th = quantile(len,0.25),
                  Median = quantile(len,0.5),
                  q75th = quantile(len,0.75),
                  Mean = mean(len),
                  SD = sd(len)) -> data
data
```

Visualize the data with box plots:
```{r,tidy=FALSE,fig.height=3,fig.width=7}
ggplot(data=ToothGrowth, aes(x=dose, y=len, fill=supp)) +
        geom_boxplot() +
        facet_grid(. ~ supp) +
        xlab("Dose [mg/day]") +
        ylab("Cell length") +
        guides(fill=guide_legend(title="Supplement type"))
```

The figure shows that the odontoblasts lenght presents a positive correlation regarding the dosage for both supplement types (VC and OJ)

##3. Use hypothesis tests to compare tooth growth by supplement type and dose

```{r,echo=FALSE}
ToothGrowth.supp_OJ_doses_.5_1 <- subset (ToothGrowth, supp == "OJ" & dose %in% c(0.5, 1.0)) 
ToothGrowth.supp_OJ_doses_.5_2 <- subset (ToothGrowth, supp == "OJ" & dose %in% c(0.5, 2.0)) 
ToothGrowth.supp_OJ_doses_1_2 <- subset (ToothGrowth, supp == "OJ" & dose %in% c(1.0, 2.0)) 
```
```{r,echo=FALSE}
t1 <- t.test(len ~ dose, data = ToothGrowth.supp_OJ_doses_.5_1,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
t2 <- t.test(len ~ dose, data = ToothGrowth.supp_OJ_doses_.5_2,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
t3 <- t.test(len ~ dose, data = ToothGrowth.supp_OJ_doses_1_2,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
results <- data.frame(c(t1$p.value,t2$p.value,t3$p.value),
                      round(c(t1$conf.int[1],t2$conf.int[1],t3$conf.int[1]),2),
                      round(c(t1$conf.int[2],t2$conf.int[2],t3$conf.int[2]),2),
                      row.names=c(".5 vs 1",".5 vs 2","1 vs 2"))
names(results) <- c("P-Value","Conf. Int. low","Conf. Int. high")
```
```{r,echo=FALSE}
kable(results,align='c',caption="Analysis of dose pairs for supplement type OJ")
```

As can be seen, all p-values are less than 0.05, therefore in all cases the null hypothesis is rejected. There is strong evidence in favour of the alternative hypothesis: *true difference in means is not equal to 0*

```{r,echo=FALSE}
ToothGrowth.supp_VC_doses_.5_1 <- subset (ToothGrowth, supp == "VC" & dose %in% c(0.5, 1.0)) 
ToothGrowth.supp_VC_doses_.5_2 <- subset (ToothGrowth, supp == "VC" & dose %in% c(0.5, 2.0)) 
ToothGrowth.supp_VC_doses_1_2 <- subset (ToothGrowth, supp == "VC" & dose %in% c(1.0, 2.0)) 
```
```{r,echo=FALSE}
t1 <- t.test(len ~ dose, data = ToothGrowth.supp_VC_doses_.5_1,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
t2 <- t.test(len ~ dose, data = ToothGrowth.supp_VC_doses_.5_2,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
t3 <- t.test(len ~ dose, data = ToothGrowth.supp_VC_doses_1_2,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
results <- data.frame(c(t1$p.value,t2$p.value,t3$p.value),
                      round(c(t1$conf.int[1],t2$conf.int[1],t3$conf.int[1]),2),
                      round(c(t1$conf.int[2],t2$conf.int[2],t3$conf.int[2]),2),
                      row.names=c(".5 vs 1",".5 vs 2","1 vs 2"))
names(results) <- c("P-Value","Conf. Int. low","Conf. Int. high")
```
```{r,echo=FALSE}
kable(results,align='c',caption="Analysis of dose pairs for supplement type VC")
```

As can be seen, all p-values are less than 0.05, therefore in all cases the null hypothesis is rejected. There is strong evidence in favour of the alternative hypothesis: *true difference in means is not equal to 0*

```{r,echo=FALSE}
ToothGrowth.dose_.5 <- subset (ToothGrowth, dose == 0.5) 
ToothGrowth.dose_1 <- subset (ToothGrowth, dose == 1) 
ToothGrowth.dose_2 <- subset (ToothGrowth, dose == 2) 
```
```{r,echo=FALSE}
t1 <- t.test(len ~ supp, data = ToothGrowth.dose_.5,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
t2 <- t.test(len ~ supp, data = ToothGrowth.dose_1,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
t3 <- t.test(len ~ supp, data = ToothGrowth.dose_2,
       paired = FALSE, var.equal = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
results <- data.frame(c(t1$p.value,t2$p.value,t3$p.value),
                      round(c(t1$conf.int[1],t2$conf.int[1],t3$conf.int[1]),2),
                      round(c(t1$conf.int[2],t2$conf.int[2],t3$conf.int[2]),2),
                      row.names=c(".5","1","2"))
names(results) <- c("P-Value","Conf. Int. low","Conf. Int. high")
```
```{r,echo=FALSE}
kable(results,align='c',caption="Analysis of supplement types OJ vs VC")
```

As can be seen, p-values are less than 0.05 for doses .5 and 1. Therefore, in this two cases the null hypothesis is rejected. There is strong evidence in favour of the alternative hypothesis: *true difference in means is not equal to 0*. On the contrary, p-value is grater than 0.05 for dose 2. Thus, we fail to reject the null hypotesis.

##4. State conclusions and the assumptions needed

**Conclusions of the study:**

1. There is strong evidence that increasing dose levels of Vitamin C leads to increased length of odontoblasts and therefore increased tooth growth in Guinea Pigs.
2. For doses of .5 and 1 mg/day, there is evidence that the administration via Oranje Juice is in favor of tooth growth
3. For doses of 2 mg/day, no difference can be satated between the two supplement types with the given data

**Main assumptions:**

1. The sample size of 60 Guinea Pigs is representative of the entire population of Guinea Pigs.
2. For the t-tests carried on, the variances are assumed to be different for the groups being compared.
3. For the t-tests carried on, the data is assumed to be unpaired.
4. The confidence level used is 95%

The code used for t test is: 
```{r,eval=FALSE}
t.test(x ~ y, data = data_subset,
       var.equal = FALSE, paired = FALSE,
       mu = 0,alternative = "two.sided",conf.level = 0.95)
```
where **x = len** and **y = dose** for *table 1* and *2*; and **y = supp** for *table 3*
