---
title: "The Exponential Distribution and the Central Limit Theorem"
author: "Luis A Sandino"
output: pdf_document
---

***
```{r,echo=FALSE,message=FALSE,results='hide'}
library(lubridate)
library(ggplot2)
Sys.setlocale("LC_TIME", "English")
```

Created on `r now()` using `r R.version$version.string`

***

## Synopsis

This project investigates the exponential distribution in R. Particularly, the distribution of averages of 40 exponentials is compared via simulation with the Central Limit Theorem (CLM). 

##Simulations

Initialization:
```{r,tidy=TRUE}
set.seed(1984) #Set the seed
lambda <- 0.2 #Set the exponential rate parameter
n <- 40 #Set the number of exponentials
sim <- 1000 #Set the number of simulations

```

Results of one simulation of 40 exponentials:
```{r,tidy=TRUE,fig.width=5,fig.height=3}
plot(rexp(n,lambda),
     main=expression(paste("One sample of 40 exponentials with rate ", lambda, " = 0.2")),
     ylab=expression(paste("exp(", lambda, ")")))
```

\clearpage

Calculate the mean and variance of 40 exponentials in 1000 simulations and plot the distribution of the mean:
```{r,tidy=TRUE}
means <- NULL
vars <- NULL
for (i in 1 : sim){
        x <- rexp(n,lambda)
        means <- c(means, mean(x))
        vars <- c(vars, var(x))
}
hist(means,col="red",breaks=40,
     main="Distribution of the mean of 40 exponentials",
     xlab="Mean")
```

##Sample Mean versus Theoretical Mean

In this section the average of sample mean is compared to the theoretical mean of the distribution. The mean of exponential distribution is $\frac{1}{\lambda}$

```{r,tidy=TRUE}
hist(means, col="blue",breaks=40,
     main="Distribution of the mean of 40 exponentials",
     xlab="Mean")
abline(v=mean(means), lwd="2", col="red")
abline(v=1/lambda, lwd="2", col="orange")
text(3.5,50,paste("Sample mean = ", round(mean(means),2)),col="red")
text(3.5,45,paste("Theoretical mean = ", 1/lambda),col="orange")
```

##Sample Variance versus Theoretical Variance

In this section the average of sample variance is compared to the theoretical variance of the distribution. The variance of exponential distribution is $\left(\frac{1}{\lambda}\right)^2$

Calculate the average of sample variance
```{r,tidy=TRUE}
print(round(mean(vars),2))
```

Calculate the theoretical variance
```{r,tidy=TRUE}
print((1/lambda)^2)
```

##Distribution

The CLT states that, the mean of a sufficiently large number of iterates of independent random variables, each with a well-defined expected value and well-defined variance, will be approximately normally distributed, regardless of the underlying distribution.

In order to check this, plot the distribution of the mean together with the distribution of $N\left(\frac{1}{\lambda},\frac{1}{\lambda\cdot\sqrt{n}}\right)$:
```{r,tidy=TRUE}
hist(means,col="blue",breaks=40,
     main="Distribution of the mean of 40 exponentials",
     xlab="Mean",freq=FALSE)
x <- seq(-3,3,length=1000)*(1/lambda) + (1/lambda)
y <- dnorm(x,1/lambda,(1/lambda)/sqrt(n))
lines(x,y,lwd=4,col="red")
text(3.5,.5,expression(paste("N",bgroup("(",paste(frac(1,lambda),",",frac(1,lambda*sqrt(n))),")"))),col="red")
```

