---
title: "Reproducible Research: Peer Assessment 1"
author: "Luis Sandino"
output: 
  html_document:
    keep_md: yes
    theme: journal
---

***
```{r,echo=FALSE,message=FALSE,results='hide'}
library(lubridate)
library(dplyr)
library(lattice)
Sys.setlocale("LC_TIME", "English")
options(scipen = 1, digits = 2)
```

Created on `r now()` using `r R.version$version.string`

***

## Loading and preprocessing the data

Read CSV file
```{r,echo = TRUE}
activity <- read.csv("activity.csv")
```

Print some data and examine data structure
```{r,echo=TRUE}
head(activity)
str(activity)
summary(activity)
```

Change date column type from factor to Date
```{r,echo=TRUE}
activity$date <- as.Date(activity$date)
```

Subset data to remove NAs
```{r,echo=TRUE}
activity.sub <- subset(activity,!is.na(activity$steps))
```

## What is mean total number of steps taken per day?

Organize data subset by date
```{r,echo=TRUE}
activity.sub <- group_by(activity.sub,date)
```

Calculate the total number of steps taken per day
```{r,echo=TRUE}
steps.per.day <- summarize(activity.sub,total=sum(steps))
print(steps.per.day)
```

Plot the total number of steps taken per day
```{r,echo=TRUE,label=Fig1}
plot(steps.per.day$total ~ steps.per.day$date,type="h",lwd=2,
     main="Total number of steps taken per day",
     xlab="Date",ylab="Total number of steps")
abline(h=mean(steps.per.day$total),lwd=2,col="green")
```

Plot the histogram of total number of steps taken per day
```{r,echo=TRUE,label=Fig2}
hist(steps.per.day$total,density=10,
     main="Histogram of total number of steps taken per day",
     xlab="Total number of steps")
```

Calculate and report the mean and median of the total number of steps taken per day
```{r,echo=TRUE}
steps.mean <- mean(steps.per.day$total)
steps.median <- median(steps.per.day$total)
```

Mean: __`r mean(steps.mean)`__

Median: __`r median(steps.median)`__

## What is the average daily activity pattern?

Organize data by time interval
```{r,echo=TRUE}
activity <- group_by(activity,interval)
```

Calculate the number of steps taken per interval, averaged across all days
```{r,echo=TRUE}
steps.per.interval <- summarize(activity,avg=mean(steps,na.rm=TRUE))
```

Look for the 5-minute interval, on average across all the days, which contains the maximum number of steps
```{r,echo=TRUE}
int.max.idx <- which.max(steps.per.interval$avg)
int.max <- steps.per.interval$interval[int.max.idx]
steps.max <- steps.per.interval$avg[int.max.idx]
```

Plot the number of steps taken per interval, averaged across all days
```{r,echo=TRUE,label=Fig3}
plot(steps.per.interval$avg ~
     steps.per.interval$interval,type="l",lwd=2,
     main="Number of steps taken per interval (averaged across days)",
     xlab="Time interval",ylab="Number of steps")
abline(v=int.max,h=steps.max,lwd=1,col="red")
```

The 5-minute interval __`r int.max`__ contains the maximum number of steps averaged across days (__`r steps.max`__)

## Imputing missing values

Calculate and report the total number of missing values in the dataset
```{r,echo=TRUE}
total.NAs <- sum(is.na(activity$steps))
per.NAs <- mean(is.na(activity$steps))
print(total.NAs)
print(per.NAs)
```

As can be seen, there are __`r total.NAs`__ NAs in the dataset which represent the __`r per.NAs*100`%__ of the data  

Create a new dataset filling in all of the missing values in the original dataset using the _interval avegarge across days_
```{r,echo=TRUE}
idx <- is.na(activity$steps)
activity.new <- activity
activity.new[idx,"steps"] <- steps.per.interval$avg[match(activity$interval[idx],steps.per.interval$interval)]
```

Organize new data by date
```{r,echo=TRUE}
activity.new <- group_by(activity.new,date)
```

Calculate the total number of steps taken per day
```{r,echo=TRUE}
steps.per.day.new <- summarize(activity.new,total=sum(steps))
print(steps.per.day.new)
```

Plot the total number of steps taken per day
```{r,echo=TRUE,label=Fig4}
par(mfrow=c(1,2),oma=c(2,1,2,0))
plot(steps.per.day$total ~ steps.per.day$date,type="h",lwd=2,
     main="Original dataset",
     xlab="Date",ylab="Total number of steps")
abline(h=mean(steps.per.day$total),lwd=2,col="green")
plot(steps.per.day.new$total ~ steps.per.day.new$date,type="h",lwd=2,
     main="New dataset with filled NAs",
     xlab="Date",ylab="")
abline(h=mean(steps.per.day.new$total),lwd=2,col="green")
title("Total number of steps taken per day", outer=TRUE)
```

As can be seen, the gaps in the original data are filled but there is no change in the mean

Plot the histogram of total number of steps taken per day
```{r,echo=TRUE,label=Fig5}
par(mfrow=c(1,2),oma=c(2,1,2,0))
hist(steps.per.day$total,density=10,
     main="Original dataset",
     xlab="Total number of steps")
hist(steps.per.day.new$total,density=10,
     main="New dataset with filled NAs",
     xlab="Total number of steps")
title("Histogram of total number of steps taken per day", outer=TRUE)
```

As can be seen, the y-axis values change a little, but the distribution of the data is quite similar

Calculate and report the mean and median of the total number of steps taken per day
```{r,echo=TRUE}
steps.mean.new <- mean(steps.per.day.new$total)
steps.median.new <- median(steps.per.day.new$total)
```

New mean: __`r mean(steps.mean.new)`__

New median: __`r median(steps.median.new)`__

As mentioned before, the mean doesn't change. The median changes and becomes equal to the mean in this case.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels ("weekday" and "weekend") indicating whether a given date is a weekday or weekend day
```{r,echo=TRUE}
period <- factor(as.numeric(wday(activity.new$date)==1|wday(activity.new$date)==7),
                 c(0,1),c("weekday","weekend"))
activity.new <- cbind(activity.new,period)
```

Organize data by time period and time interval
```{r,echo=TRUE}
activity.new <- group_by(activity.new,period,interval)
```

Calculate the number of steps taken per period and per interval, averaged across all days
```{r,echo=TRUE}
steps.per.interval.new <- summarize(activity.new,avg=mean(steps))
```

Plot for the two periods, the number of steps taken per interval, averaged across all days
```{r,echo=TRUE,label=Fig6}
xyplot(steps.per.interval.new$avg ~
     steps.per.interval.new$interval|steps.per.interval.new$period,
     type="l",lwd=2,layout=c(1,2),
     main="Number of steps taken per interval (averaged across days)",
     xlab="Time interval",ylab="Number of steps")
```

***