# Getting and Cleaning Data Course Project

## File: CodeBook.md
### Author: Luis Sandino

***

Created on `r now()` using `r R.version$version.string`

***

## Codebook

### Variables description

| Variable name             | Description                                    |
|---------------------------|------------------------------------------------|
| Subject.ID                | ID number of the subject [1-30]                |
| Activity.Name             | Activity name                                  |
| Domain.Feature.Axis.Type (Variables from 3 to 68)| where                   |
|                           | _Domain_ = time OR freq                        |
|                    |_Feature_ = mix of __words__ describing the measurement|
|                           | __Body__ -> acceleration - gravity acceleration|
|                           | __Gravity__ -> total acceleration              |
|                           | __Acc__ -> accelerometer sensor                |
|                           | __Gyro__ -> gyroscope sensor                   |
|                           | __Jerk__ -> jerk signal (from derivation)      |
|                           | __Mag__ -> Euclidean magnitude                 |
|                           | _Axis_ = X, Y OR Z direction                   |
|                           | _Type_ = mean OR std (standard deviation)      |

### Tidy Dataset structure

```{r}
str(tidyDataset)
```

### Full list of variable names

```{r}
names(tidyDataset)
```

### A few rows of data

```{r}
head(tidyDataset,10)
```

### Summary of variables

```{r}
summary(tidyDataset)
```
